Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-08-21 manuscript.tex
results edits

For purifying selection simulations, we assumed that selection coefficients followed a gamma distribution, and we considered a range of values for the mean selection coefficient against deleterious mutations ($s$) by varying the scale parameter, $k$ from \X to \X while keeping the distribution shape parameter constant (see Supporting Information for simulation commands). To make the simulation output comparable to our data, we initialized the simulations with our estimated value of $\theta$ (estimated from autosomes), and sampled 8 chromosomes per simulation from 24 populations. For simulating positive selection on the Y, we included a neutral locus of length Ln = 103 bp that is flanked by loci on either side that are subject to selection and are each of length Lf = 105 bp. Simulated sequences also contained 500 kb of linked neutral sequence from which we calculated diversity for each model. We ran 1000 replicates for each model and calculated the average $\theta_{sim}$ and the corresponding Y/A ratio of neutral diversity under the both sets of selection models separately. Finally, we tested the significance of the fit of our simulated diversity output to our empirical data (\X new figure) by calculating the proportion of simulation replicates for which $\theta_{sim}$ was not significantly different from $\theta_{observed}$, using a two-sided P value. Commands for running the simulations are provided in the Supporting Information.

@@

Our results also indicate a significant reduction in X/A diversity in the derived SC and FL sub-clades of the North Carolina race ($X/A_{FL}=0.33$ and $X/A_{SC}=0.37$) compared to the Texas race ($X/A_{TX}=0.85$) (Figure 2). Although not expected, this reduction in diversity may be associated with the recent origin of the $XY_{1}Y_{2}$ sex chromosome system, which is thought to have originated through an X-autosome fusion involving the ancestral 3rd chromosome in the Texas race \citep{smith1964evolving}. Evidence supporting this autosomal origin was recently obtained by \citep{grabowska2015}, who reported that the ancestral third chromosome in the Texas race carries the 5S rDNA locus, which is now found on both the neo-X and the $Y_{2}$ sex chromosomes in the derived North Carolina race. If recent positive selection was involved in driving the evolution of this X-A fusion, which theory suggests as likely \citep{charlesworth1980sex}, then the formerly autosomal segment on the X chromosome in the $XY_{1}Y_{2}$ sub-clades may have experienced a strong selective sweep, resulting in reduced X-linked diversity in the derived $XY_{1}Y_{2}$ sub-clades. It will be important for future work to investigate in more detail the factors driving the establishment of the X-autosome fusion in this species, and how they might impact patterns of X-linked neutral diversity.

In contrast to the situation on the X chromosome, our data indicate a strong and consistent diversity reduction on the Y chromosome, and we now consider several models that might explain this reduction.

@@

 A highly reduced level of Y chromosome variability is a predicted outcome of models Y-chromosome evolution based on the removal of deleterious mutations and linked neutral variants (e.g., background selection), but the occurrence of female-biased sex ratios in this species has also been predicted to lower Y diversity in the absence of such selective interference (\X ref). Moreover, reduced Y-linked diversity could be accented by high variance in male reproductive success (Figure 1). This is not unusual in annual plants such as \textit{R. hastatulus}, which commonly exhibit extensive phenotypic plasticity in plant size and flower production (\X Harper 1977). Conceivably, such plastic differences among plants, owing to spatial heterogeneity in resources, could influence the intensity of male-male competition, and thus affect variance in male reproductive success. Given that male plants in this wind-pollinated species produce large amounts of pollen, and that female flowers are uniovulate, strong competition among males is expected.

 In common with most flowering plants we do not have marker-based estimates of the variance in male reproductive success in \textit{R. hastatulus}. However, by comparing our empirical estimates of diversity to predictions from models that jointly predict the effects on diversity of sex ratio bias and male reproductive variance, we evaluated whether these effects could explain the level of Y/A diversity that we observed (see Methods). Conditioning on estimates of sex ratio bias in \textit{R. hastatulus} that have been estimated, ranging from $N_{m}/(N_{m}+N_{f})=0.4$ to $N_{m}/(N_{m}+N_{f})=0.35$ \citep{pickup2013influence}, the predicted Y/A diversity ratio  under a Poisson distribution of offspring numbers, is approximately 0.2 (Table1, Figure 3). Our estimated mean Y/A ratio, however, is 0.02, which is significantly lower ($\textit{P}<0.0001$), suggesting that the sex ratio effect alone is insufficient to explain our data. For larger values for the variance in male reproductive success (\X) given this sex ratio, the corresponding predictions for Y/A are all significantly higher than our observed ratio (\X table). 
 Finally, neutral models that could not be rejected, i.e., in which a highly female-biased sex ratio (\X) and a high variance in male reproductive success predict a Y/A ratio that fit out data (\X), can be excluded because they simultaneously predicted X/A ratios that were significantly different from what we observed (\X Figure). Thus, our results indicate that the combined effects of sex ratio bias and variance in reproductive success cannot jointly explain our observed levels of X, Y, and autosomal diversity in \textit{R. hastatulus}.

>>>>

For purifying selection simulations, we assumed that selection coefficients followed a gamma distribution, and we considered a range of values for the mean selection coefficient against deleterious mutations, $s$, by varying the scale parameter from 0.0001 to 0.1 while keeping the distribution shape parameter constant (see Supporting Information for simulation commands). To make the simulation output comparable to our data, we initialized the simulations with our empirically estimated autosomal $\theta$, scaled to the Y chromosome as $\theta_{Y}=\theta_{A}/4$ and adjusted for a sex ratio of $N_{m}/N_{f}=0.4$, and we sampled 8 haploid chromosomes per simulation. Simulated sequences contained 500 kb of linked neutral sequence from which we calculated diversity.

For positive selection with recurrent hitchhiking, we used sfscoder to implement the method of \citep{uricchio2014robust}, and simulated a 500kb (\X) neutral locus that was flanked by 100kb loci on either side that were subject to positive selection. We ran 1000 replicates for each model and calculated the average $\theta_{Y{sim}}$ values and the corresponding Y/A ratios for each model seperately. Finally, we tested the significance of the fit of the simulated diversity to our empirical Y/A ratio (\X new figure) by calculating the proportion of simulation replicates for which $\theta_{{Y}sim}$ was not significantly different from $\theta_{{Y}observed}$, using a two-sided P value. Commands for running the simulations are provided in the Supporting Information.

@@

Our results also indicate a significant reduction in X/A diversity in the derived SC and FL sub-clades of the North Carolina race ($X/A_{FL}=0.33$ and $X/A_{SC}=0.37$) compared to the Texas race ($X/A_{TX}=0.85$) (Figure 2). Although not expected, this reduction in diversity may be associated with the recent origin of the $XY_{1}Y_{2}$ sex chromosome system, which is thought to have originated through an X-autosome fusion involving the ancestral 3rd chromosome in the Texas race \citep{smith1964evolving}. Evidence supporting this autosomal origin was recently obtained by \citep{grabowska2015}, who reported that the ancestral third chromosome in the Texas race carries the 5S rDNA locus, which is now found on both the neo-X and the $Y_{2}$ sex chromosomes in the derived North Carolina race. If recent positive selection was involved in driving the evolution of this X-A fusion, which theory suggests can driving the evolution of such fusions \citep{charlesworth1980sex}, then the formerly autosomal segment on the X chromosome in the $XY_{1}Y_{2}$ sub-clades may have experienced a strong selective sweep, resulting in reduced X-linked diversity in the derived $XY_{1}Y_{2}$ sub-clades. It will be important for future work to investigate in more detail the factors driving the establishment of the X-autosome fusion in this species, and how they might impact patterns of X-linked neutral diversity. In contrast to the X chromosome, however, our data indicate a strong and consistent diversity reduction on the Y chromosome - an approximately 40 fold reduction compared to the mean $theta_{X}$. We next consider several possible models - neutral and selective - that might explain this reduction.

@@

 The occurrence of female-biased sex ratios in this species has been predicted to lower Y diversity due to its effects on reducing male $N_{e}$ and therefore the $N_{e}$ of the Y chromosome (\X ref). This reduction $N_{e}$ on the Y is also expected to be accented if there is high variance in male reproductive success (Figure 1), which is not unusual in annual plants such as \textit{R. hastatulus} that commonly exhibit extensive phenotypic plasticity in plant size and flower production (\X Harper 1977). Conceivably, such plastic differences, owing to spatial heterogeneity in resources, could influence the intensity of male-male competition and thus affect variance in male reproductive success. Moreover, given that male plants in this wind-pollinated species produce large amounts of pollen, and that female flowers are uniovulate, strong competition among males is expected.

 In common with most flowering plants we do not have marker-based estimates of the variance in male reproductive success in \textit{R. hastatulus}. However, by comparing our empirical estimates of diversity to predictions from models that jointly predict the effects on diversity of sex ratio bias and male reproductive variance, we evaluated whether these effects could explain the level of Y/A diversity that we observed (see Methods). Conditioning on estimates of sex ratio bias in \textit{R. hastatulus} that have been estimated, ranging from $N_{m}/(N_{m}+N_{f})=0.4$ to $N_{m}/(N_{m}+N_{f})=0.35$ \citep{pickup2013influence}, the predicted Y/A diversity ratio  is approximately 0.2 (Table1, Figure 3). This is significantly lower than our estimated mean Y/A ratio of 0.02 ($\textit{P}<0.0001$), even if we consider the upper bound from the bootstrapped or likelihood based estimated of the confidence interval. This suggests that the sex ratio effect alone is insufficient to explain our data under the standard sex ratio model.

Assuming that there is extensive variance in male reproductive success (\X), the predicted ratios of Y/A diversity (\X supp) were also significantly higher than our estimates for the empirically estimated sex ratio (\X table) . We also find, however, that purely neutral models that assumed a highly female-biased sex ratio (\X) and a level of variance in male reproductive success on the order of (\X) predicted a Y/A ratio that fit out data (\X). However, these models simultaneously predicted a range of X/A ratios that were significantly different from what we observed (\X Figure). Thus, our results indicate that the combined effects of sex ratio bias and variance in reproductive success cannot jointly explain our observed levels of X, Y, and autosomal diversity.
